Manuscript Figures

### 2021-09-17 src/manuscript_figures.Rmd
update lung cancer biomarker analysis - use the decision tree to filter candidate markers

# import assay data and subset for markers 
A <- data.table::fread(file.path(settings$assay_output$folder, 
                                 'nsclc.cnv_gex_meth_mut.csv'))
A.sub <- A[V1 %in% names(markers)]
#convert to matrix
A.subM <- as.matrix(data.frame(A.sub[,-1],
                               row.names = A.sub$V1, 
                               check.names = F))
labels <- gsub("TCGA-", "", surv[match(colnames(A.subM), bcr_patient_barcode)]$project)

# further filter top features 
top_features <- get_factor_specific_variables(t(A.subM), labels)
# get top 5 per subtype
top_markers <- get_top_by_label(top_features[padj < 0.05][order(padj)], 
                                'ref_cl', topN = 5)
# map ids 
top_markers <- merge(top_markers, map_ids(top_markers$variable, cpg2hgnc, ens2hgnc), by = 'variable')

# use treeheatr and partykit package to compute/visualize decision tree
df <- data.frame(t(round(A.subM[top_markers$variable,], 3)))
colnames(df) <- paste(top_markers$type, top_markers$hgnc, sep = ":")
df$subtype <- as.factor(labels)

set.seed(123)
inTrain <- caret::createDataPartition(df$subtype, p = 0.6)
df.train <- df[inTrain[[1]],]
df.test <- df[-inTrain[[1]],]
p5 <- treeheatr::heat_tree(
  x = partykit::ctree(subtype ~ ., data = df.train),
  data_test = df.test,
  heat_rel_height = 0.4
)

@@

p1 <- ggplot(dt[,.SD[which.min(padj)],by = c('tool', 'ref_cl')], 
       aes(x = ref_cl, y = -log10(padj))) + 
  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
  geom_hline(yintercept = -log10(0.05)) + 
  coord_flip() + 
  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
        plot.title = element_text(hjust = 0.5))

@@

dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')]
dt$tool <- toupper(gsub("_factors", "", dt$tool))
p3 <- ggplot(dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')], 
             aes(x = ref_cl, y = -log10(padj))) + 
  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
  geom_hline(yintercept = -log10(0.05)) + 
  labs(x = 'Tumor Immune Phenotypes') + 
  coord_flip() + 
  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
        plot.title = element_text(hjust = 0.5))

>>>>

# import assay data and subset for markers 
A <- data.table::fread(file.path(settings$assay_output$folder, 
                                 'nsclc.cnv_gex_meth_mut.csv'))
A.sub <- A[V1 %in% names(markers)]
#convert to matrix
A.subM <- as.matrix(data.frame(A.sub[,-1],
                               row.names = A.sub$V1, 
                               check.names = F))
labels <- gsub("TCGA-", "", surv[match(colnames(A.subM), bcr_patient_barcode)]$project)
# use treeheatr and partykit package to compute/visualize decision tree

top_markers <- map_ids(names(markers), cpg2hgnc, ens2hgnc)
# get unique table 
top_markers <- top_markers[match(top_markers[,head(variable, 1), by = c('type', 'hgnc')]$V1, variable)]

df <- data.frame(t(round(A.subM[top_markers$variable,], 3)))
colnames(df) <- paste(top_markers$type, top_markers$hgnc, sep = ":")
df$subtype <- as.factor(labels)

set.seed(123)
inTrain <- caret::createDataPartition(df$subtype, p = 0.6)
df.train <- df[inTrain[[1]],]
df.test <- df[-inTrain[[1]],]
# further filter top features: build initial tree to get subset of most important features
x = partykit::ctree(subtype ~ ., data = df.train)
selected <- names(partykit::varimp(x))
df.train <- df.train[,c(selected, 'subtype')]
df.test <- df.test[,c(selected, 'subtype')]
p5 <- treeheatr::heat_tree(
  x = partykit::ctree(subtype ~ ., data = df.train),
  data_test = df.test,
  heat_rel_height = 0.4
)

@@

p1 <- ggplot(dt[,.SD[which.min(padj)],by = c('tool', 'ref_cl')], 
       aes(x = ref_cl, y = -log10(padj))) + 
  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
  geom_hline(yintercept = -log10(0.05)) + 
  coord_flip() + 
  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
        plot.title = element_text(hjust = 0.5), 
        legend.key.size = unit(0.2, units = 'in'))

@@

dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')]
dt$tool <- toupper(gsub("_factors", "", dt$tool))
p3 <- ggplot(dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')], 
             aes(x = ref_cl, y = -log10(padj))) + 
  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
  geom_hline(yintercept = -log10(0.05)) + 
  labs(x = 'Tumor Immune Phenotypes') + 
  coord_flip() + 
  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
        plot.title = element_text(hjust = 0.5), 
        legend.key.size = unit(0.2, units = 'in'))
